---
title: "Lab 7 - k-means and PCA in R"
subtitle: "Foundations of Machine Learning (Demo), E1234"
date: "20 October 2026"
output:
  html_document:
    toc: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6.5, fig.height = 4)
set.seed(2026)
```

## The deliverable

There is no test error today, so a plot is not a result. The deliverable is **one paragraph**
(section 6) interpreting the clusters you find, plus the stability check that entitles you to
write it.

## 1. Data

Sixty fictional municipalities, three variables. Base R only - no packages beyond `stats`.

```{r data}
n <- 60
towns <- data.frame(
  town        = sprintf("T%02d", 1:n),
  pop_k       = round(c(rnorm(20, 12, 3), rnorm(20, 45, 9), rnorm(20, 160, 30)), 1),
  income_k    = round(c(rnorm(20, 28, 4), rnorm(20, 34, 5), rnorm(20, 41, 6)), 1),
  commute_min = round(c(rnorm(20, 34, 6), rnorm(20, 27, 5), rnorm(20, 21, 4)), 1)
)
head(towns)
summary(towns[, -1])
```

## 2. Standardise - not optional

k-means minimises Euclidean distance, so a variable measured in hundreds dominates one
measured in tens. Compare the two distance matrices to see the size of the effect.

```{r scale}
Z <- scale(towns[, -1])
round(head(Z, 3), 3)

d_raw    <- dist(towns[, -1])
d_scaled <- dist(Z)
cat("correlation between raw and scaled distances:",
    round(cor(as.vector(d_raw), as.vector(d_scaled)), 3), "\n")
```

## 3. Choosing k: elbow and silhouette

```{r choose-k}
wss <- sapply(2:8, function(k) kmeans(Z, k, nstart = 25)$tot.withinss)
plot(2:8, wss, type = "b", pch = 19, xlab = "k",
     ylab = "within-cluster SS", main = "Elbow plot")

# Silhouette, written out rather than imported, so nothing is hidden:
silhouette_mean <- function(Z, cluster) {
  D <- as.matrix(dist(Z))
  s <- vapply(seq_len(nrow(Z)), function(i) {
    own <- cluster == cluster[i]
    a <- if (sum(own) > 1) mean(D[i, own & seq_len(nrow(Z)) != i]) else 0
    b <- min(vapply(setdiff(unique(cluster), cluster[i]),
                    function(g) mean(D[i, cluster == g]), numeric(1)))
    (b - a) / max(a, b)
  }, numeric(1))
  mean(s)
}

for (k in 2:6) {
  km <- kmeans(Z, k, nstart = 25)
  cat(sprintf("k = %d   within-SS %8.2f   mean silhouette %.3f\n",
              k, km$tot.withinss, silhouette_mean(Z, km$cluster)))
}
```

Where is the elbow, and does the silhouette agree with it? They often do not, and that
disagreement is information: it usually means the clusters are not well separated.

## 4. Fit and profile

Cluster centres are only communicable on the original scale.

```{r fit}
km <- kmeans(Z, centers = 3, nstart = 25)
table(cluster = km$cluster)

centres_raw <- t(apply(km$centers, 1, function(z)
  z * attr(Z, "scaled:scale") + attr(Z, "scaled:center")))
round(centres_raw, 1)
```

## 5. PCA and a biplot

```{r pca}
pca <- prcomp(towns[, -1], scale. = TRUE)
summary(pca)
round(pca$rotation, 3)

plot(pca, type = "lines", main = "Scree plot")
biplot(pca, cex = 0.6, main = "PC1 vs PC2")
```

```{r pca-clusters}
scores <- pca$x[, 1:2]
plot(scores, col = km$cluster, pch = 19,
     xlab = "PC1", ylab = "PC2", main = "Clusters in PCA space")
```

## 6. Stability, then interpretation

Re-cluster on repeated 80% subsamples. If the profiles move, you found noise, not groups.

```{r stability}
agreement <- replicate(20, {
  keep <- sample(nrow(Z), round(0.8 * nrow(Z)))
  km2 <- kmeans(Z[keep, ], 3, nstart = 25)
  # Compare cluster SIZES as a crude label-invariant summary:
  paste(sort(km2$size), collapse = "-")
})
table(agreement)
```

**Write your paragraph here.** Name each cluster in substantive terms, using the profile
table from section 4 and the loadings from section 5, and state explicitly how confident the
stability check above lets you be. Three to five sentences.

## Exercises

1. **k = 2 and k = 4.** Refit and re-profile. Which of the three solutions would you present
   to a non-technical audience, and on what grounds other than the within-SS?
2. **Skip the scaling.** Re-run section 4 on `towns[, -1]` unscaled. Which variable takes
   over, and why is that a property of the metric rather than of the data?
3. **PCA first.** Cluster on `pca$x[, 1:2]` instead of on `Z`. Do the assignments change?
   Then say in one sentence why PCA can *hurt* prediction (session 5) even where it helps
   clustering.

Knit to HTML and bring your paragraph to the project clinic - the same reasoning is what the
project's "findings" section needs.
