---
title: "Lab 2 - Least squares in R"
subtitle: "Foundations of Machine Learning (Demo), E1234"
date: "15 September 2026"
output:
  html_document:
    toc: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 3.6)
set.seed(2026)
```

## Why this lab is in R

Two of the ten labs are in R, on purpose. The engine that releases these materials does not
care what language you work in, and neither should you: `lm()` and `numpy.linalg.lstsq` solve
the same problem. Knit this file to HTML and read your own output - that is the deliverable.

## 1. The data

The twelve flats from the lectures. The same rows are released as `data/housing-mini.csv`;
they are inlined here so the lab knits anywhere.

```{r data}
flats <- data.frame(
  flat_id       = sprintf("F%02d", 1:12),
  area_sqm      = c(32, 45, 52, 60, 68, 75, 80, 95, 38, 55, 110, 48),
  dist_metro_km = c(0.3, 0.9, 0.4, 1.6, 0.7, 2.1, 1.1, 0.5, 1.8, 0.6, 1.4, 2.6),
  built_year    = c(1965, 1998, 1972, 2010, 1930, 1985, 2015, 1955, 1978, 2004, 1926, 1968),
  rent_eur      = c(540, 510, 640, 545, 720, 620, 770, 860, 420, 640, 930, 400)
)
summary(flats[, -1])
```

## 2. Look before you fit

```{r explore}
plot(rent_eur ~ area_sqm, data = flats, pch = 19,
     xlab = "area (sqm)", ylab = "rent (EUR)", main = "Twelve flats")
cor(flats[, c("area_sqm", "dist_metro_km", "built_year", "rent_eur")])
```

Which pair is most strongly correlated, and which correlation is weaker than you expected?

## 3. Fit, and read the output

```{r fit}
fit <- lm(rent_eur ~ area_sqm + dist_metro_km, data = flats)
summary(fit)
```

Read three things, in this order:

1. **The coefficients.** Read off what an extra square metre is worth per month *holding
   distance fixed*, and write out that clause every time you quote the number.
2. **Residual standard error** - the model's typical miss, in euros. More interpretable than
   R-squared for anyone outside the room.
3. **R-squared** - and remember it never falls when you add a column, so it cannot choose
   between models of different size.

## 4. The same fit by hand

```{r normal-equations}
X <- cbind(intercept = 1, as.matrix(flats[, c("area_sqm", "dist_metro_km")]))
y <- flats$rent_eur
b_hat <- solve(t(X) %*% X, t(X) %*% y)   # SOLVE, do not invert
cbind(by_hand = round(as.vector(b_hat), 6), lm = round(coef(fit), 6))
```

`solve(A, b)` and `solve(A) %*% b` agree in exact arithmetic. The first is better
conditioned, and `lm()` is better still - it uses a QR decomposition.

## 5. Diagnostics

```{r diagnostics}
par(mfrow = c(1, 2))
plot(fit, which = 1)   # residuals vs fitted: curvature = a missing transform
plot(fit, which = 5)   # leverage: one far-out x can move the whole plane
par(mfrow = c(1, 1))
```

## 6. Break it deliberately

Perfect collinearity makes `X'X` singular. R does not error - it silently drops a column, and
you must notice.

```{r collinear}
flats$area_sqft <- flats$area_sqm * 10.7639   # the same variable, different units
broken <- lm(rent_eur ~ area_sqm + area_sqft + dist_metro_km, data = flats)
coef(broken)
```

Note the `NA`. Now do it the hard way and read the error:

```{r singular, error=TRUE}
Xb <- cbind(1, flats$area_sqm, flats$area_sqft)
solve(t(Xb) %*% Xb)
```

Learn the phrase "system is computationally singular" - you will meet it again, usually
because you built a dummy for every category *and* kept the intercept.

## Exercises

1. **Add `built_year`.** Does it improve the fit? Compare the residual standard error, and
   say why R-squared cannot answer this question on its own.
2. **A transform.** Fit `log(rent_eur) ~ log(area_sqm) + dist_metro_km`. How does the
   interpretation of the first coefficient change? (It becomes an elasticity.)
3. **Leverage.** Drop flat `F11` (110 sqm - the largest) and refit. How much do the
   coefficients move? Write one sentence on what that means for trusting a fit on twelve
   rows.

Knit to HTML before the next session and bring your answer to exercise 3.
